---
title: "Masters Homework 2"
author: "Iman Beckles"
date: "7/21/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Credit received ##

>- Find summary information about the total blood mercury levels in our NYC HANES data (hint: names(nyc) will call the names of each column). What is the max value, what is the minimum value.  

```{r}
nyc <- read.csv("https://raw.githubusercontent.com/fredlapolla/RVilcekMaster23/master/NYC_HANES_DIAB.csv")

names(nyc)
summary(nyc$MERCURYTOTALBLOOD)

#max value = 123.2344
#min value = 0.0919


```

>- Bonus/Challenge: What is the Standard Deviation and skewness of the blood mercury?

```{r}
MERCURYTOTALBLOODSD <- sd(nyc$MERCURYTOTALBLOOD)
print(MERCURYTOTALBLOODSD)
```



>- Install a package and load, such as "ggplot2" or "dplyr"

```{r}
#i googled the error message i got and learned that a mirror should be used when downloading packages. 
install.packages("ggplot2", repos = "https://cloud.r-project.org")
library(ggplot2)


```





>- What two commands are needed to install and use a package in R?

```{r}
#install.packages("packagename") to install once. 
#library(packagename) to load every time you use R
```




>- When you read in a .csv file into R, what class is the object that you read in (like nyc)?
#the class is a dataframe (data.frame)


>- What command can you use to determine what type of data is in a column?

```{r}
#class(object) 
#for eg: if i wanted to determine the type of data in the nyc cadmium column, I could use 
class(nyc$CADMIUM)
```


>- Provide a command to find the type of data in the DX_DBTS column of nyc. 

```{r}
class(nyc$DX_DBTS)
```


>- Provide the same command the type of data but use the other way of indexing from what you did above. 

```{r}
class(nyc[,10])

```



>- What would happen if we had chosen the first 4 columns of NYC HANES and made a matrix? What would the class of the cells in this matrix be? Why?

```{r}
class(nyc$KEY)
class(nyc$GENDER)
class(nyc$SPAGE)
class(nyc$AGEGROUP)

#the first column is character class, the following 3 are integer classes. If we make a matrix, R will coerce all the integers into characters as it there can only be 1 data type in a matrix
```


>- If we make a vector,vec<- c(3,9,4,7,8,9), and multiply it by WannaMatrix, what is the value in the third row, third column? 

```{r}
WannaMatrix <- nyc[5:10, 6:9]
WannaMatrix <- as.matrix(WannaMatrix,nrow = 6, ncol = 4)
vec <- c(3,9,4,7,8,9)
print(WannaMatrix)
NewMatrix <- vec * WannaMatrix
NewMatrix[3,3]


```



>- What happens if instead we made a vector vec2 <- c(3,9,4,7,8) and multiply it by WannaMatrix, what is the third row, third column value? What is going on? 

```{r}
vec2 <- c(3,9,4,7,8)
NewMatrix2 <- vec2 * WannaMatrix
print(NewMatrix2)
NewMatrix2[3,3]

## we get an error code because there are 5 objects in the vector but 6 in the matrix, and R cannot cleanly recycle the vector as 6 (the longer vector) is not evenly divisible by 5 (the shorter vector)
```


>- Look up the arguments for read.csv

```{r}
args(read.csv)

```


>- Use apply() to find the median value of CADMIUM, LEAD CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, and CREATININE

```{r}
LetColumns <- nyc[, c("CADMIUM", "LEAD", "CHOLESTEROLTOTAL", "GLUCOSESI", "CREATININESI", "CREATININE" )]
apply(LetColumns, 2, median, na.rm = TRUE)
```


>- Do the same using sapply()

```{r}
sapply(LetColumns, median, na.rm = TRUE)
```

>- Use a for loop to find the median value of columns 16:19

```{r}
for (i in 16:19) {
  value <- median(nyc[,i], na.rm = TRUE)
  print(c(colnames(nyc[i]), value))
}
```


>- Make a for loop to find the standard deviation (hint sd() ) of CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, AND CREATININE

```{r}
for (i in 16:19) {
  StanDev <- sd(nyc[,i], na.rm = TRUE)
  print(c(colnames(nyc[i]), StanDev))
}
```


>- Bonus/Challenge Create a vector in which to put the standard deviations 

```{r}
vec3 <- sapply(nyc[, c("CHOLESTEROLTOTAL", "GLUCOSESI", "CREATININESI", "CREATININE")], sd, na.rm = TRUE)
print(vec3)
```


>- Use ifelse to create a new variable of high and normal A1C dichotomizing with 7 as a threshold for high levels

```{r}
nyc <- na.omit(nyc)
nyc$HiA1C <- ifelse(nyc$A1C > 7 , 1, 0)
nyc$HiA1C <- factor(nyc$HiA1C)
summary(nyc$HiA1C)
```


>- Subset to only have females (gender == "female")

```{r}
#assuming 1 is male and 2 is female
females <- nyc[nyc$GENDER == 2 , ]

```


>- Create a subset of males and find the median total cholesterol level

```{r}
males <- nyc[nyc$GENDER == 1 , ]
median(males$CHOLESTEROLTOTAL)
```


>- Try to create a summary of the mean A1C using two factors, GENDER and DX_DBTS

```{r}
#by(data, INDICES, FUN)
#i can title the factors for readability
A1CMeanSummary <- by(nyc$A1C, list(Gender = nyc$GENDER, DiabetesDx = nyc$DX_DBTS), mean, na.rm = TRUE)
print(A1CMeanSummary)
```

Find the class of the first three columns of data. What are they?

```{r}
#i cant just use my sapply function becasue the A1CMeanSummary isnt a dataframe. I'm going to convert it first. 
A1CMeanSummarydf <- as.data.frame.table(A1CMeanSummary)
sapply(A1CMeanSummarydf[, 1:3], class)

```

